---
title: "The Price of Police"
author: "Robert Crump"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
rm(list = ls())

library(tidyverse)
library(treemap)
library(kableExtra)
library(scales)
library(ggthemes)
library(RColorBrewer)

library(knitr)
opts_chunk$set(echo = FALSE,
               warning = FALSE,
               message = FALSE,
               results = 'asis',
               fig.align = "center",
               fig.pos = "!H", 
               out.extra = "")

library(formatR)
opts_chunk$set(tidy.opts=list(width.cutoff=60),
               tidy=TRUE)

options(scipen = 999)

setwd("C:/Users/uberbot/Desktop/CPD/cpd_new")
```

# Introduction

2020 was a year like no other that compelled many Americans to fundamentally reconsider how their cities are policed. In Chicago, this reckoning came to a head during a weekend of protests and looting after a Minneapolis police officer murdered George Floyd on May 25th. Chicago’s mayor, Lori Lightfoot, enacted emergency powers that effectively shut down the central city while Chicago Police Department (CPD) officers fanned out through the city to quell the unrest. The city’s iconic river-spanning bridges were raised, interstate ramps were blockaded, the Chicago Transit Authority temporarily suspended service, and the Governor deployed the Illinois National Guard to deter further violence. The Summer of 2020 saw unfolding reprisals of civil unrest in different parts of the city ^[https://www.chicagotribune.com/news/breaking/ct-cb-george-floyd-fallout-impact-chicago-20201006-lkndggmch5emjiefplywhcaeou-story.html] that persisted into the Fall. As the dust settled in Chicago and elsewhere, people across the country grappled less with the “why” and more with the “how” of police reform. Police department budgets received new levels of scrutiny as many citizens and politicians openly advocated and mobilized for sweeping change.

The future of local policing depends largely on how municipal governments choose to allocate resources. Reimagining public safety and charting a new path that hopefully leads cities and the nation away from our troubled past requires a clear-eyed view of where we currently stand. The City of Chicago maintains a large Open Data Portal that contains City Budget Ordinances going back to 2011. This report examines the Fiscal Year 2021 Budget Ordinance with particular focus on the CPD budget and related appropriations. The goal of this report is to be strictly descriptive while providing a base level of knowledge for other researchers and curious citizens. However, it's worth noting that the choice of what to measure and how to present data is inherently subjective.

Summer 2021 has been a busy season for Chicago’s City Council, particularly regarding the CPD. In July, the City Council threatened a vote of no confidence in the CPD Superintendent appointed by the Mayor unless he agreed to a public grilling on community violence ^[https://chicago.suntimes.com/city-hall/2021/6/30/22557658/chicago-violence-shootings-fourth-july-weekend-police-plans-brown-council-aldermen-special-meeting]. Weeks later, the same Council passed the Mayor’s proposal for a citizen oversight board that she originally crafted as the President of the Police Board ^[https://blockclubchicago.org/2021/07/21/city-council-approves-first-ever-civilian-led-commission-to-oversee-chicago-police-department/
]. The recently negotiated contract between the City and the Fraternal Order of Police (CPD union) will substantially increase the CPD budget while also adopting new accountability measures recommended by the Department of Justice ^[https://news.wttw.com/2021/07/26/police-officers-get-back-pay-face-new-accountability-rules-under-proposed-8-year-deal
]. On July 30th, the Civilian Office of Police Account ability began searching for a new Chief Administrator ^[https://www.chicagocopa.org/press/the-city-of-chicago-begins-official-search-and-selection-process-for-the-next-chief-administrator-of-copa/]. Police spending in Chicago is certain to go up over the coming years, and how that money is spent will remain a contentiously argued topic.

What may come of these new changes is hard to predict for the beleaguered Department, and for the City. A shared understanding of facts is the first step towards meaningful change, and finding a path forward, at the very least, requires a compass and a map. I used two sources for this initial report, both from the City of Chicago Open Data Portal. The primary resource is the Fiscal Year 2021 Budget Ordinance ^[https://data.cityofchicago.org/Administration-Finance/Budget-2021-Budget-Ordinance-Appropriations/6tbx-h7y2], and the secondary resource is the Fiscal Year 2021 Budget Ordinance – Positions and Salaries ^[https://data.cityofchicago.org/Administration-Finance/Budget-2021-Budget-Ordinance-Positions-and-Salarie/gcwx-xm5a] which is a detailed subset of the full budget ordinance.


```{r load data}
budget_ordinances <- read_csv("data/budget_ordinances.csv")

FY21 <- budget_ordinances %>% 
  filter(year == 2021)

titles_salaries <- read_csv("data/titles_salaries.csv")

ts_FY21 <- titles_salaries %>% 
  filter(year == 2021,
         position_control == "1")
```

# Chicago City 2021 Budget Overview

The City of Chicago, like most local governments in the United States, divides its budget into three broad categories: the general fund, enterprise funds, and special revenue funds. The City of Chicago calls its general fund the Corporate Fund, which contains most of the CPD budget. Enterprise and special revenue funds are highly restricted in discretionary spending choices because they require stability to support essential services. The largest enterprise fund for most US municipalities is the water and/or sewer funds. Special revenue funds, such as fiduciary funds (pensions), are also legally structured to resist changes in spending choices between election cycles. Although municipal general funds permit more flexibility, most of the budget is similarly constrained under contract, such as, for example, the Chicago FOP contract. The motivating goal of constrained budgeting is stability and predictability of public investment. Moreover, the dynamics of political economy compels negotiators to demand credible commitments from politicians who may or may not be in office for the duration of a contract.

When engaging with something like the City of Chicago budget, the first noticeable characteristics are its size and complexity. Through close study and data manipulation using R, I’ve attempted to display these characteristics while illustrating my decisions for the sake of interpretation and gathering insight by narrowing focus on particular datapoints. To start, let’s get a picture of the proportions of every named City fund using a treemap chart.

```{r full budget treemap}
FY21 %>% 
  group_by(fund_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>%
  treemap(
    treemap_2021,
    index = c("fund_description"),
    vSize = "total_app",
    type = "index",
    title = "Chicago City Budget Ordinance 2021, ~13.5 Billion",
    palette = "GnBu",
    fontsize.labels = c(10),
    fontcolor.labels = c("black"),
    bg.labels=c("transparent"),
    align.labels = list(c("left", "top")),
    overlap.labels = 0,
    inflate.labels = F,
    border.lwds = 1)
```

The chart above gives us a superficial level of insight, but mostly demonstrates the number of individual funds within the City budget; 251 in total. To aid legibility and increase information, I grouped all the funds below the top ten largest into a ‘Miscellaneous’ category. I also shortened a few names and divided the appropriation amount by one million so I could fit the dollar amount within the same square region.

```{r reduced budget treemap}
top_ten_FY21 <- FY21 %>% 
  group_by(fund_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>% 
  arrange(desc(total_app)) %>% 
  head(10) %>% 
  rename("top_ten" = fund_description)

FY21 %>% 
  
  # label top ten city funds and lump all the rest in 'other category'
  mutate(
    fund_description = case_when(
      !fund_description %in% top_ten_FY21$top_ten ~ "MISCELLANEOUS",
      
      # relabel for sizing
      fund_description == "MUNICIPAL ANNUNITY AND BENEFIT" ~ "MUNI PENSION FUND",
      fund_description == "FIREMEN'S ANNUITY AND BENEFIT" ~ "CFD PENSION",
      fund_description == "MIDWAY AIRPORT FUND" ~ "MIDWAY",
      fund_description == "HEAD START AND EARLY HEAD START" ~ "HEAD START",
      TRUE ~ fund_description
      )
    ) %>% 
  group_by(fund_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>%
  mutate(total_app = round(total_app / 1000000),
         
         # use numeric variable for sizing and character for label
         money = paste0("$", total_app)) %>% 
  treemap(
        index = c("fund_description", "money"),
        vSize = "total_app",
        type = "index",
        title = "Chicago Budget by Fund, Millions of $",
        palette = "GnBu",
        fontcolor.labels = c("black", "black"),
        fontsize.labels = c(10),
        bg.labels=c("transparent"),
        align.labels = list(c("left", "top"),
                          c("right", "bottom")),
        overlap.labels = 0,
        inflate.labels = F,
        border.lwds = 1)
```

In tabular format, we can include more exact numbers and add a percentage column.

```{r reduced budget table}
FY21 %>% 
  mutate(fund_description = if_else(
    !fund_description %in% top_ten_FY21$top_ten,
      "MISCELLANEOUS",
      fund_description
    )
  ) %>% 
  group_by(fund_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>%
  mutate(perc_total = round((total_app/sum(total_app))*100, 2)) %>% 
  arrange(desc(total_app)) %>% 
  kable(align = "lccrr",
        format.args = list(big.mark = ","),
        caption = "Chicago Budget Appropriations by Fund",
        col.names = c("Fund Description", "Total Appropriation ($)", "% of Total"),
        "simple", 
        table.placement = "H")
```

The Corporate Fund comprises about 30% of the total City budget at \$4.03 billion. The proliferation of enterprise and special revenue funds makes Miscellaneous the second biggest category. 33 different departments get at least a portion of their budget from the Corporate Fund, of which CPD is largest by a significant proportion at \$1.55 billion or 38.56% of the total.

```{r reduced corporate fund treemap}
corp_top_ten <- FY21 %>% 
  filter(fund_description == "CORPORATE FUND") %>% 
  group_by(department_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>% 
  arrange(desc(total_app)) %>% 
  head(10) %>% 
  rename("top_ten" = department_description)

FY21 %>% 
  filter(fund_description == "CORPORATE FUND") %>% 
  mutate(department_description = if_else(
    !department_description %in% corp_top_ten$top_ten,
    "MISCELLANEOUS",
    department_description
    )
  ) %>% 
  group_by(department_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>% 
  mutate(total_app = round(total_app / 1000000),
         money = paste0("$", total_app)) %>% 
  treemap(
      index = c("department_description", "money"),
      vSize = "total_app",
      type = "index",
      title = "Chicago Corporate Fund Appropriations, Millions of $",
      palette = "GnBu",
      fontsize.labels = c(10, 10),
      fontcolor.labels = c("black", "black"),
      bg.labels=c("transparent"),
      align.labels = list(c("left", "top"),
                        c("right", "bottom")),
      inflate.labels = F)
```

```{r corp fund table}
FY21 %>% 
  filter(fund_description == "CORPORATE FUND") %>% 
  mutate(department_description = if_else(
    !department_description %in% corp_top_ten$top_ten,
    "MISCELLANEOUS",
    department_description
    )
  ) %>% 
  group_by(department_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>% 
  mutate(perc_total = round((total_app/sum(total_app))*100, 2)) %>%
  arrange(desc(total_app)) %>% 
  kable(align = "lccrr",
        format.args = list(big.mark = ","),
        caption = "Corporate Fund by Department",
        col.names = c("Dept Description", "Total Appropriation ($)", "% of Total"),
        "simple", 
        table.placement = "H")
```

Next, let’s take a quick look at Positions and Salary data for the City by department. One important note about the Positions and Salary dataset is the ‘position control’ variable that designates whether a position is hourly or salaried. For our purposes, hourly positions need to be filtered out because it affects how the position counts and appropriations are counted. Salaried positions are counted individually, and appropriations specify that annual salary for each position. Hourly positions are counted in hours per position not number of positions, and appropriations specify how much money is available to payout at the hourly rate for a position. CPD hourly positions (mostly trainees) account for roughly $10 million.

```{r ts by dept table}
ts_top_ten_pos <- ts_FY21 %>% 
  group_by(department_description) %>% 
  summarize(total_app = sum(total_budgeted_unit)) %>% 
  rename("top_ten" = department_description) %>% 
  arrange(desc(total_app)) %>% 
  head(10)

ts_FY21 %>%
  mutate(department_description = if_else(
    !department_description %in% ts_top_ten_pos$top_ten,
    "MISCELLANEOUS",
    department_description
    )
  ) %>% 
  group_by(department_description) %>% 
  summarize(emply_num = sum(total_budgeted_unit),
            sal_app = sum(total_budgeted_amount)) %>% 
  mutate(perc_total = round((emply_num/sum(emply_num))*100, 2),
         perc_total_sal = round((sal_app/sum(sal_app))*100, 2)) %>%
  select(department_description,
         emply_num,
         perc_total,
         sal_app,
         perc_total_sal) %>% 
  arrange(desc(emply_num)) %>% 
  kable(align = "lccrr",
        format.args = list(big.mark = ","),
        caption = "Positions by Department",
        col.names = c("Department", 
                      "Total Positions", 
                      "% of Positions", 
                      "Salary App. ($)",
                      "% Salary App."),
        "simple", 
        table.placement = "H")
```

The CPD employs 42% of the City’s total workforce and salary appropriations. For a visual comparison of positions by department, let’s use a horizontal bar chart.

```{r positions by dept bar chart}
ts_FY21 %>% 
  group_by(department_description) %>% 
  summarize(emply_num = sum(total_budgeted_unit)) %>% 
  ggplot() +
  geom_col(aes(
    fct_reorder(department_description, emply_num), 
    emply_num, 
    fill = emply_num)
    ) +
  labs(title = "Number of Salaried Positions by Department") +
  scale_fill_continuous_tableau() +
  theme_fivethirtyeight() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8)) +
  coord_flip()
```

# CPD Budget Analysis

The chart below shows a simplified list of the appropriations for the CPD in the 2021 budget. For the sake of legibility, I filtered out any items below \$2 million divided the amounts by \$1 million. CPD is primarily a payroll expense, however, there are a few line items outside of payroll that represent different forms of compensation. We’ll explore appropriation categories after a quick detour into the CPD Positions and Salaries data.

```{r CPD appropriations by expense type}
FY21 %>% 
  filter(department_number == "57") %>% 
  
  # "expense type"
  group_by(appropriation_account_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>%
  
  # for better visibility, we're filtering out less than 2M
  filter(total_app >= 2000000) %>%
  mutate(total_app = round(total_app / 1000000)) %>% 
  ggplot() +
  geom_col(aes(
    fct_reorder(appropriation_account_description, total_app), 
    total_app,
    fill = total_app)
    ) +
  labs(title = "CPD 2021 Budget Appropriations",
       subtitle = "Millions of Dollars") +
  scale_y_continuous(labels = dollar_format()) +
  scale_fill_continuous_tableau() +
  theme_fivethirtyeight() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8)) +
  coord_flip()
```

I decided to leave this particular table unreduced to fully display a meaningful layer of complexity. There are 114 job titles within the CPD, although many are different types of assignments for police officers.

```{r CPD appropriations by job title}
# start with long list of titles. position control == "1" indicates salary position.
# hourly positions account for about 10M

ts_FY21 %>% 
  filter(department_code == 57 &
         position_control == "1") %>% 
  group_by(title_description) %>% 
  summarize(total_app = sum(total_budgeted_amount),
            total_employed = sum(total_budgeted_unit)) %>% 
  arrange(desc(total_app)) %>% 
  kable(align = "lccrr",
        format.args = list(big.mark = ","),
        caption = "CPD Appropriation by Job Title",
        col.names = c("Title Description", "Total Appropriation ($)", "Number of Employees"),
        "simple", 
        table.placement = "H")
# we see a lot of police officers assigned with varying titles
```

The primary distinction between CPD job titles is between peace officers and non-peace officers as defined by under Illinois statute (720 ILCS 5/2-13 ^[https://www.ilga.gov/legislation/ilcs/fulltext.asp?DocName=072000050K2-13]). The Office of the Inspector General (OIG) maintains a list of job title codes that signify peace officer designation ^[https://informationportal.igchicago.org/cpd-sworn-officer-unit-assignments/]. I used this list of job title codes as a reference applied to job title codes in the Positions and Salaries dataset to divide CPD employees into either category. However, several police officer assignments and the Deputy Superintendent were not included in the OIG list. Using basic knowledge and inductive reasoning, I applied the peace officer designation to all relevant job title codes and descriptions.

```{r appropriation by PO status}
# define peace officer title code:: https://informationportal.igchicago.org/glossary/sworn-officer/
PO_title_codes <- c("9015", "9016", "9126", "9151", "9152", "9153", "9154", "9155", "9158", "9160", "9161", "9164", "9165", "9168", "9171", "9173", "9174", "9175", "9201", "9206", "9752", "9781", "9785", "9796", "9957")

ts_FY21 %>% 
  filter(department_code == 57 &
         position_control == "1") %>% 
  
  # label peace officers by title codes & title description
  mutate(
    peace_officer = case_when(
      title_code %in% PO_title_codes ~ "Peace Officer",
      str_detect(title_description, "Police Officer") ~ "Peace Officer",
      str_detect(title_description, "Deputy Superintendent") ~ "Peace Officer",
      TRUE ~ "Non Peace Officer")
    ) %>% 
  group_by(peace_officer) %>%
  summarize(total_app = sum(total_budgeted_amount)/1000000) %>% 
  ggplot() +
  geom_col(aes(peace_officer, total_app), 
           fill = "skyblue",
           alpha = 0.7) +
  labs(title = "Salary Appropriation by Job Type",
       subtitle = "Millions of $") +
  scale_y_continuous(labels = dollar_format()) +
  theme_fivethirtyeight() +
  theme(panel.grid.major.x = element_blank())
```

Although the budget appropriation for peace officer salaries is much higher than non-peace officers, their average annual salary ranges are fairly close. This is the broadest categorization possible, so future research could reveal more about how the hierarchy among peace officers affects compensation costs.

```{r PO vs. non-PO salary range}
ts_FY21 %>% 
  filter(department_code == 57 &
         position_control == "1") %>% 
  
  # label peace officers by title codes & title description
  mutate(peace_officer = case_when(title_code %in% PO_title_codes ~ "Peace Officer",
                                   str_detect(title_description, "Police Officer") ~ "Peace Officer",
                                   str_detect(title_description, "Deputy Superintendent") ~ "Peace Officer",
                                   TRUE ~ "Non Peace Officer")) %>% 
  group_by(peace_officer) %>%
  summarise(av_sal = weighted.mean(budgeted_pay_rate, total_budgeted_unit),
            
            # need to use this package to calculate weighted standard deviation
            sd = sqrt(Hmisc::wtd.var(budgeted_pay_rate, total_budgeted_unit))) %>%  
  ggplot() +
  geom_col(aes(peace_officer, av_sal), 
           fill = "skyblue",
           alpha = 0.7) +
  geom_errorbar(aes(x = peace_officer, ymin = av_sal - sd, ymax = av_sal + sd),
                width = 0.4, 
                colour = "orange", 
                alpha = 0.9, 
                size = 1.3) +
  labs(title = "Annual Salary Range by Job type") +
  scale_y_continuous(labels = dollar_format()) +
  theme_fivethirtyeight() +
  theme(panel.grid.major.x = element_blank())

# hm. so non POs make on average significantly lesss, but the confidence intervals overlap
```

# Total Policing Costs

The price of police is more than the CPD budget line in the City government. Most notably, the Policemen’s Annuity and Benefit Fund (pension) is an actuarial fund completely separate from the Corporate Fund that falls under the authority of the City Department on Finance. State law mandates the level of yearly City contributions to the pension fund, which are set to increase until the fund reaches actuarial solvency. Other meaningful costs is the Civilian Office of Police Accountability, the Police Board, and various Consent Decree ^[http://chicagopoliceconsentdecree.org/] appropriations scattered across different departments. Below is a table of all the appropriations either contained within the CPD budget specifically or ancillary to it. This list is almost certainly not exhaustive, but based on what I can surmise for myself, the grand total for policing in Chicago for Fiscal Year 2021 is \$2.53 billion.

```{r CPD plus ancillary costs}
# just like the titles list, this is obviously unwieldy and un-interpret-able
FY21 %>% 
  filter(fund_code == "0683" | # policeman's annuity fund
         department_number == "55" | # police department
         department_number == "57" | # police board
         department_number == "60" | # copa
         appropriation_account_description == "CONSENT DECREE") %>% # misc consent decree
  group_by(appropriation_account_description) %>% 
  summarise(total_app = sum(ordinance_amount)) %>%
  arrange(desc(total_app)) %>% 
  kable(align = "lccrr",
        format.args = list(big.mark = ","),
        caption = "Total Costs",
        col.names = c("Appropriation Type", "Total Appropriation ($)"),
        "simple", 
        table.placement = "H")
```

```{r reduced typology}
medical_costs <- c(
                   "HOS/MED-NOT WORKER COMP",
                   "HOSPITAL//MEDICAL - NOT WORKERS' COMP"
                   )

compensation_other <- c(
                        "COMPENSATORY TIME PAYMENT",
                        "CONTRACT WAGE-PR",
                        "CONTRACT WAGE-SALARY",
                        "CONTRACT WAGE - SALARY",
                        "CROSSING GUARDS - VACATION PAYOUT",
                        "DUTY AVAILABILITY",
                        "FOR REIMB OTHER FUNDS",
                        "FRINGE BENEFITS",
                        "FURLOUGH / COMP TIME BUY-BACK",
                        "FURLOUGH/COMP TIME",
                        "LESS SALARY SAVINGS FROM UNPAID TIME OFF",
                        "PAYMENT RETROACTIVE SALARIES",
                        "REIMBURSABLE OVERTIME",
                        "SAL AND WAGES-ON VOUCHER",
                        "SALARIES AND WAGES - ON VOUCHER",
                        "SALARY PROVISION",
                        "SCHEDULE SALARY ADJ",
                        "SCHEDULE SALARY ADJUSTMENTS",
                        "SCHEDULED WAGE ADJUSTMENTS",
                        "SPECIALTY PAY",
                        "STIPENDS",
                        "SUPERVISORS QUARTERLY PAYMENT",
                        "SWORN/CIVILIAN HOLIDAY PREMIUM PAY",
                        "SWORN/CIVILIAN PAY",
                        "TUITION REIMBURSEMENT"
                        )

physical_assets <- c(
                     "APPARATUS AND INSTRUMENTS",
                     "BOOKS AND RELATED MATERIAL",
                     "BOOKS AND RELATED MTL",
                     "CLEANING AND SAN SUPPLY",
                     "CLEANING AND SANITATION SUPPLIES",
                     "CLOTHING",
                     "COMMUNICATION DEVICES",
                     "DATA CIRCUITS",
                     "DRUGS  MEDICINE AND CHEMICAL MATERIALS",
                     "DRUGS MED AND CHEM MATL S",
                     "EQUIPMENT",
                     "FOOD",
                     "FREIGHT AND EXPRESS CHARGES",
                     "FREIGHT/EXPENSE CHARGES",
                     "GASOLINE",
                     "IT MAINTENANCE",
                     "LEASE/PURCHASE EQUIPMENT",
                     "LICENSE STICKER TAG AND PLATES",
                     "LIVESTOCK",
                     "MACHINERY AND EQUIPMENT",
                     "MAINT FACILITIES",
                     "MAINTENANCE AND OPERATION - VEHICLES",
                     "MAINTN & OPERTN-CY VEHL",
                     "MATERIAL AND SUPPLIES",
                     "MOBILE COMM SERVICES",
                     "MOBILE COMMUNICATION SERVICES",
                     "POSTAGE",
                     "PUBLICATN & REPRODT-OUT",
                     "PUR VEHICLES-POL/FIRE",
                     "PURCHASE-DATA HARDWARE",
                     "PURCHASE - DATA HARDWARE",
                     "RENTAL-DATA HARDWARE EQ",
                     "RENTAL EQUIPMENT AND SERVICES",
                     "REPAIR PARTS AND MATERIAL",
                     "REPAIR/MAINT EQUIPMENT",
                     "STATIONERY AND OFFICE SUPPL",
                     "STATIONERY AND OFFICE SUPPLIES",
                     "TECH AND SCIENTIFIC EQUIP",
                     "TECHNICAL AND SCIENTIFIC EQUIPMENT",
                     "TECHNICAL MEETING COSTS",
                     "TELEPHONE-CENTREX",
                     "TELEPHONE-MAINTENANCE",
                     "TELEPHONE - CENTREX BILLINGS",
                     "TELEPHONE - MAINTENANCE",
                     "TELEPHONE - NON-CENTREX BILLINGS",
                     "TELEPHONE NON-CENTREX",
                     "TOOL <= $100 UNIT",
                     "TOOLS > $100 UNIT",
                     "UNIFORM ALLOWANCE",
                     "VEHICLES",
                     "VEHICLES-POLICE",
                     "VEHICLES - POLICE AND FIRE"
                     )

prof_services <- c(
                   "ADVERTISING",
                   "COURT REPORTING",
                   "GRAPHIC DESIGN SERV",
                   "OFF AND BUILDING SERV",
                   "OFFICE AND BUILDING SERVICES",
                   "OUTSIDE GRAPHIC SERVICES",
                   "PROF & TECHNICAL SERVICES",
                   "PROF SERV-IT MAINT",
                   "PROFESSIONAL AND TECHNICAL SERVICES",
                   "SOFTWARE MAINT/LICENSING",
                   "SOFTWARE MAINTENANCE AND LICENSING",
                   "STATISTICAL STUDIES",
                   "TESTING AND INSPECTING",
                   "WASTE DISPOSAL SERV",
                   "WASTE DISPOSAL SERVICES"
                   )

misc_appropriations <- c(
                         "CONTINGENCIES",
                         "CULTURAL PROGRAMMING GRANTS",
                         "DELEGATE AGENCIES",
                         "DUES  SUBSC & MEM",
                         "INDIRECT COSTS",
                         "LEGAL EXPENSES",
                         "LOCAL TRANSPORTATION",
                         "LOSS IN COLLECTION OF TAXES",
                         "MATCHING AND SUPPLEMENTAL GRANTS",
                         "PHYSICAL EXAMS",
                         "REIMBURSABLE TRANSFERS BETWEEN FUNDS",
                         "REIMBURSEMENT - 2FM",
                         "REIMBURSEMENT - AIS",
                         "REIMBURSEMENT - CPD",
                         "REIMBURSEMENT - DAIS",
                         "REIMBURSEMENT - DSS",
                         "REIMBURSEMENT TO TRAVELERS",
                         "REIMBURSEMENT TRAVELERS",
                         "REIMREIMB CORP",
                         "TRANSFER FOR SERVICES - DSS",
                         "TRANSPORTATION AND EXPENSE ALLOWANCE",
                         "VIOLENCE REDUCTION PROGRAM"
                         )
```

To simplify this table, I created four categories for medical costs, non-salary compensation, physical assets, and professional services.

```{r total price by category reduced}
# so let's sort into categories of our own choosing and view with horizontal bar chart

FY21 %>% 
  filter(fund_code == "0683" |
         department_number == "55" |
         department_number == "55" | # police department
         department_number == "57" | # police board
         department_number == "60" | # copa
         appropriation_account_description == "CONSENT DECREE") %>% # misc consent decree
  mutate(
      appropriation_account_description = case_when(
        department_description == "COPA" ~ "COPA",
        appropriation_account_description %in% compensation_other ~ "COMPENSATION - OTHER",
        appropriation_account_description %in% medical_costs ~ "MEDICAL COSTS",
        appropriation_account_description %in% prof_services ~ "PROFESSIONAL SERVICES",
        appropriation_account_description %in% physical_assets ~ "PHYSICAL ASSETS",
        appropriation_account_description %in% misc_appropriations ~ "MISCELLANEOUS",
        TRUE ~ appropriation_account_description)
  ) %>% 
  group_by(appropriation_account_description) %>% 
  summarize(total_app = sum(ordinance_amount)) %>% 
  mutate(total_app = round(total_app / 1000000)) %>% 
  arrange(desc(total_app)) %>%  
  ggplot() +
  geom_col(aes(
    fct_reorder(appropriation_account_description, total_app), 
    total_app, 
    fill = total_app)
           ) +
  labs(title = "Total Price by Category",
       subtitle = "Millions of $") +
  scale_y_continuous(labels = dollar_format()) +
  scale_fill_continuous_tableau() +
  theme_fivethirtyeight() +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank()) +
  coord_flip()
```

# Conclusion

The total Chicago Police Department Budget Appropriation (including appropriations outside the Corporate Fund) is \$1.6 billion dollars for Fiscal Year 2021, or 12.56% of the full City budget. It's proportion of the Corporate Fund is \$1.5 billion or 38.56%. Most City employees work for the CPD (~14 thousand) and the overwhelming majority of salary expenses flow towards that department. However, the budget line for CPD does not capture all the direct costs of policing for the City. Pension contributions (\$818 million), the Civilian Office of Police Accountability, the Police Board, and various Consent Decree expenses bring the total up to \$2.5 billion.

Opportunities for future research include a closer look at personnel costs and how compensation is structured through the CPD hierarchy. The City also maintains datasets on crime and COPA cases, which could prove insightful with regard to effectiveness and performance measures. These data along with past City budgets can shed light on what has come before, where we are now, and what the future may hold. There is also a vast literature that can enhance my qualitative analysis that I am eager to reengage.